Finally, functional class scoring enrichment analysis of Infinium DNA methylation data

### 2023-10-30 manuscript/manuscript.Rmd
adding simulation to the ms

Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
This is made more complicated for Infinium array data due to the presence of multiple probes per gene,
meaning the data needs to be summarised from probes to genes to gene sets.

We propose three different approaches for applying enrichment analysis depending on when the differential
methylation step is conducted (**Figure 1**).

1. <strong>Differential methylation analysis is conducted on probes and these results are aggregated
to the gene and gene set level (called "LA").<.strong>
In this approach differential analysis was conduced on probe values using limma.
Differential probe methylation results (t-statistics) were then aggregated to the gene level by median.
These median t-statistics were subject to downstream enrichment analysis using a one-sample two-sided
t-test (a self-contained test) to estimate whether the t-statistic distribution of the gene set varies
from zero.
The use of a self-contained test is warranted as global hyper- or hypo-methylation is common between
conditions, where competitive tests may give misleading results.
The use of parametric tests is also warranted, as probe t-statistics typically exhibit a normal
distribution, even in the presence of global hyper- or hypo-methylation.

2. <strong>Differential methylation analysis conducted after probe-level methylation is aggregated to
gene-level (called "AL").</strong>
As each gene may have many probes, these individual data points were summarised into a single
value for each individual.
We used the mean of M-values for each probe belonging to a gene, which generated a new gene-level
M-value matrix.
This gene-level methylation matrix underwent differential analysis using limma, followed by an
enrichment test as above (one-sample two-tailed t-test).

3. <strong>Methylation values are aggregated to the gene set level for differential testing
(called "AA").</strong>
M-values for each probe were aggregated to the gene level by mean, and then aggregated again to the gene
set level by mean or median.
The resulting matrix of aggregated M-values underwent differential analysis with limma and the results
including log fold changes and significance values were used to infer pathway regulation.
This approach has a caveat in that it may partly violate the assumption of independent sampling, as
it is known that some genes are represented in multiple gene sets.

@@

>>>>

Functional enrichment analysis is a process of data summarisation from genes to gene sets (pathways).
This is made more complicated for Infinium array data due to the presence of multiple probes per gene,
meaning the data needs to be summarised from probes to genes and then to gene sets.
Here, we outline seven potential approaches for FCS of Infinium methylation data:

1. Limma Average t-test (LAT).
Differential methylation analysis is conducted at the probe level with limma, and the limma
t-statistics for each gene are summarised (mean).
The mean t-statistics are used in a downstream two-sample two-way t-test of gene set enrichment.

2. Limma Top t-test (LTT).
As above, except instead of calculating the mean t-statistic, the probe with the largest magnitude is selected
to represent the gene.

3. Limma Average Wilcox (LAW).
Similar to LAT, except instead of the two-sided t-test of gene set enrichment, a non-parametric alternative,
the Wilcoxon signed-rank test.

4. Limma Average Mitch (LAM).
Similar to LAW, however instead of the Wilcoxon test, the Mitch package is used to execute a two-way,
two-sample ANOVA-on-ranks test (Kruskal Wallis).

5. Aggregate Limma t-test (ALT).
In this method, all probe measurements beloning to each gene are averaged before conducting differential
methylation andlysis with limma.
The limma t-statistics are then used downstream for enrichment analysis with the two-sample two-way t-test.

6. Aggregate Limma Wilcox (ALW).
Similar to ALT, however it uses the Wilcoxon non-parametric test for gene set enrichment.

7. Aggregate Limma Mitch (ALM).
Similar to ALW, however uses the Mitch package to execute a two-way, two-sample ANOVA-on-ranks test
(Kruskal Wallis).

The above were compared to a standard ORA based approach (GSA), which involves limma on probes, selection of
statistically significant probes (FDR<0.05) for ORA with the `GSAmeth()` function of missMethyl which conducts a
modified hypergeometric test that accounts for multiple probe biases.
In the case that fewer than 250 significant probes are identified, then 250 probes with the smallest p-values
were selected.

@@

To assess these methods, we adopted a simulated data approach.
EPIC Infinium array data derived from non-cancerous lung tissue of 37 cancer patients was used for our
evaluations.
We randomly sampled datasets to serve as control and cases (no replacement), with group sizes varying between
3 and 12.
One thousand random gene sets were created with sizes varying between 20 and 100 members.
Randow gene sets were used to avoid problems caused by the large overlap among real gene sets.
Throughout our evaluations, 50 gene sets were selected to be differentially methylated, 25 with increased and 25
with decreased methylation.
From these gene sets, half of the members were selected.
For those selected genes, half of the annotated probes were selected.
We adjusted the m-values in the case group by a specified amount, which we call the "delta", which we varied
between 0.1 and 0.4.
Following the incorporation of methylation changes to selected probes in the case group, the data underwent
enrichment analysis using `GSAmeth` approach, or one of the seven FCS methods outlined above.
Enrichment results were then filtered for FDR<0.05 and correct direction of methylation change.
Differentially methylated gene sets observed were compared to gene sets selected to be the ground truth, so that
the number of true positives (TP), false positives (FP) and false negatives (FN) could be determined.
At each setting for group size, delta and gene set size, 20 replications were conducted each with a set seed
to ensure reproducibility.
Mean TP, FP and FN values were determined, and then precision, recall and F1 scores were calculated.
